---
title: "Assignment 5"
author: "Qingyuan Pei [Jump to Github](https://github.com/MiaPei88/BINF6210-Assignement5)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction

Molecular Phylogenetics and Congruence

### Description of Data Set

The data set I choose for this project is the DNA sequences of COI gene and CytB gene in *Acipenser* from NCBI, and this data set is obtained on December 3, 2023. COI and CytB genes are widely used as phylogenetic markers of fish(Ivanova et al., 2007; Li et al., 2018). Here I use rentrez package and functions in Entrez_Functions.R, a course material, to obtain the data in this R script. I also limited the sequence lengths of the two genes to exclude whole genome sequences: the range of sequence length is 500-700 for COI gene and is 600-1200 for CytB gene. 

### Code Section 1 - – Data Acquisition, Exploration, Filtering, and Quality Control

Loading packages in this project.
```{r, message = FALSE}
library(tidyverse)
library(rentrez)
library(Biostrings)
library(ggplot2)
library(DECIPHER)
library(ape)
```

Search for the hits of COI and CytB gene of *Acipenser*.
```{r}
NCBI_search <- entrez_search(
  db = "nuccore", 
  term = "Acipenser[ORGN] AND COI[Gene] AND 500:700[SLEN]", 
  use_history = T
  )
NCBI_search

NCBI_search1 <- entrez_search(
  db = "nuccore", 
  term = "Acipenser[ORGN] AND CytB[Gene] AND 600:1200[SLEN]", 
  use_history = T
  )
NCBI_search1
```

Using functions in Entrez_Functions.R to obtain data set from NCBI.
```{r, results = "hide"}
# Save the functions wrote in the R script to the current environment
source("Entrez_Functions.R")
# Fetch fasta files from NCBI
FetchFastaFiles(searchTerm = "Acipenser[ORGN] AND COI[Gene] AND 500:700[SLEN]", 
                seqsPerFile = 100, fastaFileName = "Acipenser_fetch_1.fasta")
FetchFastaFiles(searchTerm = "Acipenser[ORGN] AND CytB[Gene] AND 600:1200[SLEN]", 
                seqsPerFile = 100, fastaFileName = "Acipenser_fetch_2.fasta")
# Merge all fasta files and store it as a data frame 
dfNCBI_COI <- MergeFastaFiles(filePattern = "Acipenser_fetch_1.fasta.*")
dfNCBI_CytB <- MergeFastaFiles(filePattern = "Acipenser_fetch_2.fasta.*")

# Add two new columns to include the species name and sequence length of each row,
# and filter out whole genome sequences and sequences with more than 5% of Ns.
Filter1 <- function(df) {
  filtered1 <- df %>%
    mutate(Species_Name = word(Title, 2L, 3L)) %>%
    mutate(Sequence_Length = nchar(Sequence)) %>%
    mutate(Gene_Type = word(str_extract(Title, "(?i)partial|complete"))) %>%
    filter(Gene_Type == "partial") %>%
    filter(str_count(Sequence, "N") <= (0.05 * Sequence_Length))
  return(filtered1)
}
dfCOI <- Filter1(dfNCBI_COI) 
dfCytB <- Filter1(dfNCBI_CytB) 
```

Draw a histogram to see the distribution of COI sequence lengths to further decide the following data filtering steps.
```{r, fig.align = 'center', fig.cap = '**Figure 1**: '}
hist((dfCOI$Sequence_Length), 
     xlab = "Sequence Length", 
     ylab =  "Frequency", 
     main = "Distribution of COI Sequence Lengths")
```

Draw a histogram to see the distribution of CytB sequence lengths to further decide the following data filtering steps.
```{r, fig.align = 'center', fig.cap = '**Figure 2**: '}
hist((dfCytB$Sequence_Length), 
     xlab = "Sequence Length", 
     ylab =  "Frequency", 
     main = "Distribution of CytB Sequence Lengths")
```

Filter out sequences that their length are not the majority.
```{r}
dfCOI_filtered <- dfCOI %>%
  #remove sequences with length shorter than 600
  filter(Sequence_Length >= 600)

dfCytB_filtered <- dfCytB %>%
  #remove sequences that are longer and shorter than the median legnth +/- 100
  filter(Sequence_Length >= median((Sequence_Length) - 100) & 
           Sequence_Length <= median((Sequence_Length) + 100))
```

Check if the number of species between 10 to 50.
```{r}
length(unique(dfCOI$Species_Name))
length(unique(dfCytB$Species_Name))
intersect(dfCOI$Species_Name,dfCytB$Species_Name)
```


```{r, results='hide'}
AlignmentTest <- function(filtered_df) {
  df <- as.data.frame(filtered_df)
  df$nucleotides <- DNAStringSet(df$Sequence)
  df$Alignment_id <- paste(word(df$Species_Name, 2L), 
                                  word(df$Title, 1L),
                                  sep = " ")
  names(df$nucleotides) <- df$Alignment_id
  alignment.test <- AlignSeqs(df$nucleotides)
  return(alignment.test)
}

dfCOI.alignment.test <- AlignmentTest(dfCOI_filtered)
BrowseSeqs(dfCOI.alignment.test)
```

ON097839.1, MT455724.1, MK903722.1, MK903723.1, MK903724.1, MK903725.1, MK903726.1, HM902200.1, MF122058.1, MF122059.1, KC578836.1, KC578837.1, KF929572.1, KF558304.1, KC500133.1, KC578827.1, EU523886.1, EU523878.1, KM286430.1, JQ623904.1, JQ623905.1, JQ623906.1, KX145481.1, KX145032.1, KC500105.1 


```{r, results='hide'}
dfCOI_filtered1 <- dfCOI_filtered %>%
  filter(
    !grepl('ON097839.1|MT455724.1|MK903722.1|MK903723.1|MK903724.1|MK903725.1|MK903726.1|HM902200.1|MF122058.1|MF122059.1|KC578836.1|KC578837.1|KF929572.1|KF558304.1|KC500133.1|KC578827.1|EU523886.1|EU523878.1|KM286430.1|JQ623904.1|JQ623905.1|JQ623906.1|KX145481.1|KX145032.1|KC500105.1', Title))

AlignTranslationTest <- function(filtered_df) {
  df <- as.data.frame(filtered_df)
  df$nucleotides <- DNAStringSet(df$Sequence)
  df$Alignment_id <- paste(word(df$Species_Name, 2L), 
                                  word(df$Title, 1L),
                                  sep = " ")
  names(df$nucleotides) <- df$Alignment_id
  align.translation.test <- AlignTranslation(df$nucleotides, 
                                             readingFrame=NA,
                                             type="AAStringSet")
  return(align.translation.test)
}

dfCOI.translation <- AlignTranslationTest(dfCOI_filtered1)

BrowseSeqs(dfCOI.translation)
```

ON097800.1, ON097762.1, ON097679.1, MT455123.1, KX145047.1, MG648396.1, MG648380.1, MG648371.1, MG648367.1, MG648366.1, KX145047.1, GQ328798.1, GQ328790.1

```{r, results='hide'}
dfCOI_final <- dfCOI_filtered1 %>%
  filter(
    !grepl('ON097800.1|ON097762.1|ON097679.1|MT455123.1|KX145047.1|MG648396.1|MG648380.1|MG648371.1|MG648367.1|MG648366.1|KX145047.1|GQ328798.1|GQ328790.1', Title))

DendrogramCheck <- function(filtered_df, clustering_method, clustering_threshold) {
  df <- as.data.frame(filtered_df)
  df$nucleotides <- DNAStringSet(df$Sequence)
  df$Alignment_id <- paste(word(df$Title, 1L))
  names(df$nucleotides) <- df$Alignment_id
  df <- AlignSeqs(df$nucleotides)
  dnaBin <- as.DNAbin(df)
  distanceMatrix <- dist.dna(dnaBin, model = "TN93", 
                           as.matrix = TRUE, pairwise.deletion = TRUE)
  clusters <- DECIPHER::TreeLine(myDistMatrix = distanceMatrix,
                                      method = clustering_method,
                                      cutoff = clustering_threshold,
                                      showPlot = TRUE,
                                      type = "both",
                                      verbose = TRUE)
  return(clusters)
}

clusters.COI <- DendrogramCheck(dfCOI_final, "NJ", 0.012)
```


```{r}
unique(dfCOI_final$Species_Name)
```


```{r, results='hide'}
dfCytB.alignment.test <- AlignmentTest(dfCytB_filtered)
BrowseSeqs(dfCytB.alignment.test)
```

FJ974042.1, AJ249693.1, AJ249694.1, FJ974041.

```{r, results='hide'}
dfCytB_filtered1 <- dfCytB_filtered %>%
  filter(
    !grepl('FJ974042.1|AJ249693.1|AJ249694.1|FJ974041.', Title))

dfCytB.translation <- AlignTranslationTest(dfCytB_filtered1)

BrowseSeqs(dfCytB.translation)
```

FJ974040.1, AJ277594.1, FJ974043.1

```{r, results='hide'}
dfCytB_final <- dfCytB_filtered1 %>%
  filter(
    !grepl('FJ974040.1|AJ277594.1|FJ974043.1', Title))

clusters.CytB <- DendrogramCheck(dfCytB_final, "NJ", 0.015)
```


```{r}
unique(dfCytB_final$Species_Name)
```


### Main Software Tools Description


### Code Section 2 - Main Analysis

```{r}

```


### Results and Discussion


### Reflection


### Acknowledgements


### References

Ivanova, N. V., Zemlak, T. S., Hanner, R. H., & Hebert, P. D. (2007). Universal primer cocktails for fish DNA barcoding. *Molecular Ecology Notes*, 7(4), 544-548.

Li, X., Shen, X., Chen, X., Xiang, D., Murphy, R. W., & Shen, Y. (2018). Detection of potential problematic Cytb gene sequences of fishes in GenBank. *Frontiers in genetics*, 9, 30.