---
title: "Assignment 5"
author: "Qingyuan Pei"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction

Molecular Phylogenetics and Congruence

### Description of Data Set

The data set I choose for this project is the DNA sequences of COI gene and CytB gene in *Acipenser* from NCBI, and this data set is obtained on December 3, 2023. COI and CytB genes are widely used as phylogenetic markers of fish(Ivanova et al., 2007; Li et al., 2018). Here I use rentrez package and functions in Entrez_Functions.R, a course material, to obtain the data in this R script. I also limited the sequence lengths of the two genes to exclude whole genome sequences: the range of sequence length is 500-700 for COI gene and is 600-1200 for CytB gene. 

### Code Section 1 - – Data Acquisition, Exploration, Filtering, and Quality Control

Loading packages in this project.
```{r, message = FALSE}
#tidyverse version: 2.0.0
library(tidyverse)
library(rentrez)
library(Biostrings)
library(ggplot2)
```

Search for the hits of COI and CytB gene of *Acipenser*.
```{r}
NCBI_search <- entrez_search(
  db = "nuccore", 
  term = "Acipenser[ORGN] AND COI[Gene] AND 500:700[SLEN]", 
  use_history = T
  )
NCBI_search

NCBI_search1 <- entrez_search(
  db = "nuccore", 
  term = "Acipenser[ORGN] AND CytB[Gene] AND 600:1200[SLEN]", 
  use_history = T
  )
NCBI_search1
```
Using functions in Entrez_Functions.R to obtain data set from NCBI.
```{r, results = "hide"}
# Save the functions wrote in the R script to the current environment
source("Entrez_Functions.R")
# Fetch fasta files from NCBI
FetchFastaFiles(searchTerm = "Acipenser[ORGN] AND COI[Gene] AND 500:700[SLEN]", 
                seqsPerFile = 100, fastaFileName = "Acipenser_fetch_1.fasta")
FetchFastaFiles(searchTerm = "Acipenser[ORGN] AND CytB[Gene] AND 600:1200[SLEN]", 
                seqsPerFile = 100, fastaFileName = "Acipenser_fetch_2.fasta")
# Merge all fasta files and store it as a data frame 
dfNCBI_COI <- MergeFastaFiles(filePattern = "Acipenser_fetch_1.fasta.*")
dfNCBI_CytB <- MergeFastaFiles(filePattern = "Acipenser_fetch_2.fasta.*")

# Add two new columns to include the species name and sequence length of each row,
# and filter out whole genome sequences and sequences with more than 5% of Ns.
Filter1 <- function(df) {
  filtered1 <- df %>%
    mutate(Species_Name = word(Title, 2L, 3L)) %>%
    mutate(Sequence_Length = nchar(Sequence)) %>%
    mutate(Gene_Type = word(str_extract(Title, "(?i)partial|complete"))) %>%
    filter(Gene_Type == "partial") %>%
    filter(str_count(Sequence, "N") <= (0.05 * Sequence_Length))
  return(filtered1)
}
dfCOI <- Filter1(dfNCBI_COI) 
dfCytB <- Filter1(dfNCBI_CytB) 
```

Draw a histogram to see the distribution of COI sequence lengths to further decide the following data filtering steps.
```{r, fig.align = 'center', fig.cap = '**Figure 1**: '}
hist((dfCOI$Sequence_Length), 
     xlab = "Sequence Length", 
     ylab =  "Frequency", 
     main = "Distribution of COI Sequence Lengths")
```

Draw a histogram to see the distribution of CytB sequence lengths to further decide the following data filtering steps.
```{r, fig.align = 'center', fig.cap = '**Figure 2**: '}
hist((dfCytB$Sequence_Length), 
     xlab = "Sequence Length", 
     ylab =  "Frequency", 
     main = "Distribution of CytB Sequence Lengths")
```

Filter out sequences that their length are not the majority.
```{r}
dfCOI_filtered <- dfCOI %>%
  #remove sequences with length shorter than 600
  filter(Sequence_Length >= 600)

dfCytB_filtered <- dfCytB %>%
  #remove sequences that are longer and shorter than the median legnth +/- 100
  filter(Sequence_Length >= median((Sequence_Length) - 100) & 
           Sequence_Length <= median((Sequence_Length) + 100))
```

Check if the number of species between 10 to 50.
```{r}
length(unique(dfCOI$Species_Name))
length(unique(dfCytB$Species_Name))
```

### Main Software Tools Description


### Code Section 2 - Main Analysis

```{r}

```


### Results and Discussion


### Reflection


### Acknowledgements


### References


Ivanova, N. V., Zemlak, T. S., Hanner, R. H., & Hebert, P. D. (2007). Universal primer cocktails for fish DNA barcoding. *Molecular Ecology Notes*, 7(4), 544-548.

Li, X., Shen, X., Chen, X., Xiang, D., Murphy, R. W., & Shen, Y. (2018). Detection of potential problematic Cytb gene sequences of fishes in GenBank. *Frontiers in genetics*, 9, 30.